Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-11-26 Scripts/Manuscript.Rmd
small details

Here, we estimate the time of emergence and threat point of the share distribution of transboundary stocks worldwide. We employ a species distribution model driven by output from an Earth system model with ten ensemble members to project the distribution of the 663 selected species shared [@PalaciosAbrantes:2020tv] by 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario RCP8.5 (see Methods). We treated each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population [@PalaciosAbrantes:2020tv] and only considered shared stocks between neighboring EEZs (i.e., transboundary stocks, excluding the high seas in this study). Our results suggest that many transboundary stocks have already ‘emerged’ from historical natural variability and overpassed their share threat point in most neighboring EEZs, posing a threat to the sustainability of shared stocks and the resilience of international treaties.

@@

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@PalaciosAbrantes:2020tv]; @Teh:2015gd], resulting in a total of 9,132 transboundary stocks in this analysis. Using the transboundary species Atlantic cod (Gadus morhua) as an example, based on the definition of transboundary stock in this study, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share another stock of G. morhua, but the United States and Greenland do not share a stock of this transboundary species. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided into Mexico Pacific and Mexico Atlantic; \@ref(fig:toeFigureS1))) including island territories. We determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by the geopolitical regions according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and by natural biome [@Reygondeau:2019uh]. The habitat association of each species was determined based on information available from FishBase for fish species (http://www.fishbase.org) and SeaLifeBase for invertebrates (http://www.fishbase.org). For each stock and EEZ, we used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g., sea surface temperature) whereas demersal species distribution is driven by ocean bottom variables (e.g., bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (see Cheung et al. [-@Cheung:2010dt; -@Cheung:2016jd] for model details). DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by Exclusive Economic Zones that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd], Cheung et al. *in revision*].

We forced the DBEM with simulated ocean conditions from a ten-ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the ‘high’ climate change emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. In the GFDL-ESM2M, atmospheric surface temperature is projected to increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a stocks’ distribution, we have to understand both the historical as well as the future climate-driven distribution variability, to infer differences between time frames. We do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M-ensembles were started from infinitesimally small differences in Earth system initial conditions resulting in a unique atmosphere and ocean state at each point in time. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

@@

We developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:toeFigureS2). The centroid of a transboundary stock was determined by the average (µ) latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top 95th percentile (e.g., per = 95%). We ran a sensitivity analysis using a subset of species (n = 34) for all EEZs (per = 20th, 50th and $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:toeFigureS3)). The centroid of each EEZ was estimated using the st package in R (Figure \@ref(fig:toeFigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, we computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

@@

where $lat_{eez}$ and $lat_{ts}$ are the latitudes of the EEZ and transboundary stock centroids, respectively and $lon_{eez}$ and $lon_{ts}$ are the respective longitudes. Then, for each year we calculated the transboundary index as follows:

@@

Knowing the point in time (e.g., year) at which the distribution of a shared stock will diverge from its natural internal variability is important in informing the lead-time to which climate adaptation needs to be implemented [@Link:2010ei]. Here, we define time of emergence as the time when the stock shared between neighbouring EEZs will exceed the historical natural internal variability as follows:

@@

where $\mu_{TI}$ is the yearly mean TI ($\mu$) across all ten ensemble members in year *yr* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical reference period (1951 - 2005). We set a threshold of TI above which we consider a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., Equation \@ref(eq:four), representing a probability of 68% that the transboundary index has emerged from historical variability). We tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., "threat point" [@Sumaila:2013hv]. Here, we define threat point as the minimum proportion of a shared stock within an EEZ that a country will accept to engage in cooperative management with the country’s sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the shared stock.

First we estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. We did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, we average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($SSR_{th}$) spans from 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each stock. We then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 ($SR{te}$), and the mid $21^{st}$ century as the average of 2041 to 2060 ($SSR{tm}$). We chose these time periods to match the challenges of achieving fisheries-related UN-SDGs (6) such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we defined a threat point for each EEZ’s stock as $SSR_{th} \pm \sigma$, where $\sigma$ is the standard deviation of $SSR_{th}$. This way, a change of SSR beyond an EEZ’s threat point happens when the future SSR exceeds one standard deviation of the historical variations of the SSR i.e when $SSR_{et} \geqq (SSR_{th} + \sigma)$ or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, we estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_{th}$) for each stock whose share ratio overpassed the threat point following Palacios-Abrantes et al. (*in press*),

@@

We tested the results for normality (e.g. skewness = 1.02, kurtosis = 2.83) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat associations, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software R version 3.5.2 (2018-12-20; Eggshell Igloo) with the packages data.table [@Packagedatatable:2019uh], janitor [@PackagejanitorSim:2018ur], rfishbase [@PackagerfishbaseR:2019th], sf [@PackagesfSimpleF:2018vp], sp [@PackagespClasses:2019tn], tidiverse [@PackagetidyverseE:2017vq], tidytext [@PackagetidytextTe:2019vk], gmt [@PackagegmtInterfa:2017vh] and zoo [@PackagezooSInfr:2019tm], ggrepel [@PackageggrepelAut:2020to], zeallot [@PackagezeallotMul:2018wy], viridis [@PackageviridisDef:2018us], cowplot [@CowplotStreamlined:2019wt], moments [@PackagemomensMome:2015uh] and pgirmess [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

The average time of emergence of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 \pm 26 years (Figure \@ref(fig:toeFigure1)). In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. Fish stocks with transboundary index that first emerged from their historical variability in each EEZ was projected to be as early as 2006 and the last one in 2100. About 55% (n = 3,154) of those emerging stocks are projected to have their transboundary index emerged between 2006 and 2020 (e.g. Peruvean anchoveta - *Engraulis ringens* - shared by Chile and Peru emerged in 2010). The number of EEZs with emerging stocks is projected to increase steeply from 87% in 2006 to 96% in 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.7% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)). Moreover, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 and around one fifth (19%) of the EEZs have an average time of emergence of the transboundary index across stocks prior to 2020. Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century (Figure Figure \@ref(fig:toeFigure1)). We based our transboundary index on a transboundary stock’s centroid defined as the average grid cells that present the higher stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that our method of estimating the time of emergence of the transboundary index is robust to the abundance level (e.g. the top 90%) of the stock (Figure \@ref(fig:toeFigureS3).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence by Exclusive Economic Zone (EEZ) and transboundary stock: A) results in absolute numbers (green). B) results expressed as a percentage of the total emerging stocks and EEZ with emerging stock (purple)."}

@@

The median time of emergence of the transboundary index of stocks varies significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, DF = 93, p < 0.001; Figure \@ref(fig:toeFigure2)A). Overall, most tropical EEZs will see an earlier time of emergence with the EEZs of Latin America and the Caribbean and Polynesia having significantly earlier times of emergence (p < 0.05; see Table \@ref(tab:toeTableS1)) for test statistics) than almost all other regions e.g., Ecuador (ToE = 2013) and Papa New Guinea (ToE = 2018) (Figure \@ref(fig:toeFigure2)B). In contrast, EEZs located in template regions like northern Europe and eastern Asia will have a significantly later time of emergence (p < 0.05; see Table \@ref(tab:toeTableS1 for test statistics) than the rest of the world, e.g., Ireland (ToE = 2067) and North Korea (ToE = 2058). However, notable exceptions of tropical EEZs with a later average time of emergence across stocks include Pacific EEZ of Honduras (ToE = 2071) and Panama (ToE = 2070). Moreover, despite being outside the tropics, Belgium (ToE = 2015) and Norway (ToE = 2019) will have relatively early average time of emergence (Figure \@ref(fig:toeFigure2)A). Only in a few cases, the transboundary index of a stock will not emerge between 2006 and 2100. In the Arctic EEZs of Canada and the US, the present study included three stocks: capelin (*Mallotus villosus*), saffron cod (*Eleginus gracilis*), and Pacific herring (*Clupea pallasii*). In the case of Brazil, we included four transboundary stocks, all shared with Uruguay [@PalaciosAbrantes:2020tv]: Argentine menhaden (*Brevoortia pectinata*), Argentine shortfin squid (*Illex argentinus*), Argentine shrimp (*Pleoticus muelleri*), and pink shrimp (*Penaeus paulensis*). In all cases, our study suggest that the transboundary index of these stocks will not emerge before 2100 relative to the historic (1951-2005) distribution (Figure \@ref(fig:toeFigure2)A)

@@

On average, stocks that will see their transboundary index emerge by 2100 represent 27% \pm 23% of their EEZ’s revenue produced by fisheries targeting transboundary species between 2004 and 2010. However, large variation exists in the proportion that stocks with emerging transboundary index have in the current revenue participation across EEZs. In some cases e.g., El Salvador and New Zealand, stock’s with emerging transboundary index represent less than 1% of the revenue generated from transboundary fisheries  in their respective EEZs between 2004 and 2014. On the contrary, stocks with emerging transboundary index from the Marshal Islands and Tokelau represent over 90% of the revenue generated from fisheries targeting transboundary stocks within their respective EEZs during the same time period (Figure \@ref(fig:toeFigure2)A). In addition, some EEZs are projectede to have a few emerging stocks that represent large proportion of revenue from transboundary stocks e.g., in Peru and Finland (with 10 and 5 emerging stocks representing 86% and 78% of average fishing revenue between 2004 to 2010, respectively). Morocco and Malaysia are the only two countries amongst the top 10 countries with the larger number of stocks with emerging transboundary index (n = 56 and 53, respectively) where the contribution of such stocks on fishing revenue is less than 20%.

@@

We set an arbitrary threshold of 1 s.d. (67% confidence) to determine the time of emergence of the transboundary index and the change in stock share ratio below the threat point (see Methods). We tested the sensitivity of our methods by considering a larger threshold for the time of emergence (e.g., 2 s.d for a 95% confidence). The global average time of emergence of the transboundary index considering a 2 s.d threshold was 2036 \pm 28, a difference of about 7 years relative to the 1 s.d. threshold (Figure \@ref(fig:toeFigureS2)). In total, 4,119 transboundary stocks would emerge between 2006 and 2100 considering a more conservative metric, that is, 40% (n = 1626) fewer emerging stocks than using a 1.s.d threshold.

>>>>

Here, I estimate the time of emergence and threat point of the share distribution of transboundary stocks worldwide. I employ a species distribution model driven by output from an Earth system model with ten ensemble members to project the distribution of the 663 selected species shared [@PalaciosAbrantes:2020tv] by 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario RCP8.5 (see Methods). I treated each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population [@PalaciosAbrantes:2020tv] and only considered shared stocks between neighboring EEZs (i.e., transboundary stocks, excluding the high seas in this study). Our results suggest that many transboundary stocks have already ‘emerged’ from historical natural variability and overpassed their share threat point in most neighboring EEZs, posing a threat to the sustainability of shared stocks and the resilience of international treaties.

@@

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@PalaciosAbrantes:2020tv; @Teh:2015gd], resulting in a total of 9,132 transboundary stocks in this analysis. Using the transboundary species Atlantic cod (Gadus morhua) as an example, based on the definition of transboundary stock in this study, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share another stock of G. morhua, but the United States and Greenland do not share a stock of this transboundary species. The boundaries of the world’s EEZs were defined using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided into Mexico Pacific and Mexico Atlantic; Figure \@ref(fig:toeFigureS1))) including island territories. The intersections between polygons were defined using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by the geopolitical regions according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and by natural biome [@Reygondeau:2019uh]. The habitat association of each species was determined based on information available from FishBase for fish species (http://www.fishbase.org) and SeaLifeBase for invertebrates (http://www.fishbase.org). For each stock and EEZ, I used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. I report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

The distribution of marine species were proejcted using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g., sea surface temperature) whereas demersal species distribution is driven by ocean bottom variables (e.g., bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (see Cheung et al. [-@Cheung:2010dt; -@Cheung:2016jd] for model details). DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by Exclusive Economic Zones that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd, Cheung et al. *in revision*].

The DBEM was forced with simulated ocean conditions from a ten-ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the ‘high’ climate change emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. In the GFDL-ESM2M, atmospheric surface temperature is projected to increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a stocks’ distribution, I have to understand both the historical as well as the future climate-driven distribution variability, to infer differences between time frames. I do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M-ensembles were started from infinitesimally small differences in Earth system initial conditions resulting in a unique atmosphere and ocean state at each point in time. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

@@

I developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:toeFigureS2)). The centroid of a transboundary stock was determined by the average (µ) latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, I included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top 95th percentile (e.g., per = 95%). I ran a sensitivity analysis using a subset of species (n = 34) for all EEZs (per = 20th, 50th and $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:toeFigureS3)). The centroid of each EEZ was estimated using the st package in R (Figure \@ref(fig:toeFigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, the distance between centroids was computed assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

@@

where $lat_{eez}$ and $lat_{ts}$ are the latitudes of the EEZ and transboundary stock centroids, respectively and $lon_{eez}$ and $lon_{ts}$ are the respective longitudes. Then, for each year I calculated the transboundary index as follows:

@@

Knowing the point in time (e.g., year) at which the distribution of a shared stock will diverge from its natural internal variability is important in informing the lead-time to which climate adaptation needs to be implemented [@Link:2010ei]. Here, I define time of emergence as the time when the stock shared between neighbouring EEZs will exceed the historical natural internal variability as follows:

@@

where $\mu_{TI}$ is the yearly mean TI ($\mu$) across all ten ensemble members in year *yr* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical reference period (1951 - 2005). I set a threshold of TI above which it was consider a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., Equation \@ref(eq:four), representing a probability of 68% that the transboundary index has emerged from historical variability). I tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., "threat point" [@Sumaila:2013hv]. Here, I define threat point as the minimum proportion of a shared stock within an EEZ that a country will accept to engage in cooperative management with the country’s sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the shared stock.

First I estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. I did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, I average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($SSR_{th}$) spans from 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each stock. I then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 ($SR{te}$), and the mid $21^{st}$ century as the average of 2041 to 2060 ($SSR{tm}$). I chose these time periods to match the challenges of achieving fisheries-related UN-SDGs (6) such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, I defined a threat point for each EEZ’s stock as $SSR_{th} \pm \sigma$, where $\sigma$ is the standard deviation of $SSR_{th}$. This way, a change of SSR beyond an EEZ’s threat point happens when the future SSR exceeds one standard deviation of the historical variations of the SSR i.e when $SSR_{et} \geqq (SSR_{th} + \sigma)$ or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, I estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_{th}$) for each stock whose share ratio overpassed the threat point following Palacios-Abrantes et al. (*in press*),

@@

The time of emergence results were tested for normality (e.g. skewness = 1.02, kurtosis = 2.83) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, I tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of the transboundary index of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat associations, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software R version 3.5.2 (2018-12-20; Eggshell Igloo) with the packages data.table [@Packagedatatable:2019uh], janitor [@PackagejanitorSim:2018ur], rfishbase [@PackagerfishbaseR:2019th], sf [@PackagesfSimpleF:2018vp], sp [@PackagespClasses:2019tn], tidiverse [@PackagetidyverseE:2017vq], tidytext [@PackagetidytextTe:2019vk], gmt [@PackagegmtInterfa:2017vh] and zoo [@PackagezooSInfr:2019tm], ggrepel [@PackageggrepelAut:2020to], zeallot [@PackagezeallotMul:2018wy], viridis [@PackageviridisDef:2018us], cowplot [@CowplotStreamlined:2019wt], moments [@PackagemomensMome:2015uh] and pgirmess [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

The average time of emergence of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:toeFigure1)). In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. Fish stocks with transboundary index that first emerged from their historical variability in each EEZ was projected to be as early as 2006 and the last one in 2100. About 55% (n = 3,154) of those emerging stocks are projected to have their transboundary index emerged between 2006 and 2020 (e.g. Peruvean anchoveta - *Engraulis ringens* - shared by Chile and Peru emerged in 2010). The number of EEZs with emerging stocks is projected to increase steeply from 87% in 2006 to 96% in 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.7% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)). Moreover, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 and around one fifth (19%) of the EEZs have an average time of emergence of the transboundary index across stocks prior to 2020. Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century (Figure \@ref(fig:toeFigure1)). I based our transboundary index on a transboundary stock’s centroid defined as the average grid cells that present the higher stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that our method of estimating the time of emergence of the transboundary index is robust to the abundance level (e.g. the top 90%) of the stock (Figure \@ref(fig:toeFigureS3)).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence by Exclusive Economic Zone (EEZ) and transboundary stock."}

@@

The median time of emergence of the transboundary index of stocks varies significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, DF = 93, p < 0.001; Figure \@ref(fig:toeFigure2)A). Overall, most tropical EEZs will see an earlier time of emergence with the EEZs of Latin America and the Caribbean and Polynesia having significantly earlier times of emergence (p < 0.05; see Table \@ref(tab:toeTableS1)) for test statistics) than almost all other regions e.g., Ecuador (ToE = 2013) and Papa New Guinea (ToE = 2018) (Figure \@ref(fig:toeFigure2)B). In contrast, EEZs located in template regions like northern Europe and eastern Asia will have a significantly later time of emergence (p < 0.05; see Table \@ref(tab:toeTableS1 for test statistics) than the rest of the world, e.g., Ireland (ToE = 2067) and North Korea (ToE = 2058). However, notable exceptions of tropical EEZs with a later average time of emergence across stocks include Pacific EEZ of Honduras (ToE = 2071) and Panama (ToE = 2070). Moreover, despite being outside the tropics, Belgium (ToE = 2015) and Norway (ToE = 2019) will have relatively early average time of emergence (Figure \@ref(fig:toeFigure2)A). Only in a few cases, the transboundary index of a stock will not emerge between 2006 and 2100. In the Arctic EEZs of Canada and the US, the present study included three stocks: capelin (*Mallotus villosus*), saffron cod (*Eleginus gracilis*), and Pacific herring (*Clupea pallasii*). In the case of Brazil, there were four transboundary stocks included, all shared with Uruguay [@PalaciosAbrantes:2020tv]: Argentine menhaden (*Brevoortia pectinata*), Argentine shortfin squid (*Illex argentinus*), Argentine shrimp (*Pleoticus muelleri*), and pink shrimp (*Penaeus paulensis*). In all cases, our study suggest that the transboundary index of these stocks will not emerge before 2100 relative to the historic (1951-2005) distribution (Figure \@ref(fig:toeFigure2)A)

@@

On average, stocks that will see their transboundary index emerge by 2100 represent 27% $\pm$ 23% of their EEZ’s revenue produced by fisheries targeting transboundary species between 2004 and 2010. However, large variation exists in the proportion that stocks with emerging transboundary index have in the current revenue participation across EEZs. In some cases e.g., El Salvador and New Zealand, stock’s with emerging transboundary index represent less than 1% of the revenue generated from transboundary fisheries  in their respective EEZs between 2004 and 2014. On the contrary, stocks with emerging transboundary index from the Marshal Islands and Tokelau represent over 90% of the revenue generated from fisheries targeting transboundary stocks within their respective EEZs during the same time period (Figure \@ref(fig:toeFigure2)A). In addition, some EEZs are projectede to have a few emerging stocks that represent large proportion of revenue from transboundary stocks e.g., in Peru and Finland (with 10 and 5 emerging stocks representing 86% and 78% of average fishing revenue between 2004 to 2010, respectively). Morocco and Malaysia are the only two countries amongst the top 10 countries with the larger number of stocks with emerging transboundary index (n = 56 and 53, respectively) where the contribution of such stocks on fishing revenue is less than 20%.

@@

I set an arbitrary threshold of 1 s.d. (67% confidence) to determine the time of emergence of the transboundary index and the change in stock share ratio below the threat point (see Methods). I tested the sensitivity of our methods by considering a larger threshold for the time of emergence (e.g., 2 s.d for a 95% confidence). The global average time of emergence of the transboundary index considering a 2 s.d threshold was 2036 $\pm$ 28, a difference of about 7 years relative to the 1 s.d. threshold (Figure \@ref(fig:toeFigureS2)). In total, 4,119 transboundary stocks would emerge between 2006 and 2100 considering a more conservative metric, that is, 40% (n = 1626) fewer emerging stocks than using a 1.s.d threshold.

I compared species time of emergence within EEZs by species’ habitat association (Figure \@ref(fig:toeFigure4)). Each species is assigned to be associated to one of the seven habitat types based on information available from FishBase and SeaLifeBase for fish and invertebrates, respectively (see Methods). I found significant differences in the time of emergence of stocks range shifts, indicated by the transbounudary index, by their specie's habitat association (Kruskal-Wallis, $X^2$ = 286.48, DF = 93, p < 0.001). Specifically, time of emergence of range shifts of species that are associated with pelagic-oceanic and reef habitats were significantly different than other habitat preferences (See Table \@ref(tab:toeTableS2)). On average, species associated with pelagic oceanic habitats e,g, tunas (*Thunnus sp.*), bathy-pelagic habitats e.g., blue whiting (*Micromesistius poutassou*) and corals (e.g., Bigeye snapper *Lutjanus lutjanus*) are projected to have an earliest time of emergence (toeFigure \@ref(fig:toeFigure4)). Species associated with bathydemersal habitats such as Alaska plaice (*Pleuronectes quadrituberculatus*) generally have an average time of emergence of range shifts that is later than 2025 (Figure \@ref(fig:toeFigure4)).
